---
title: "How a Model Learns in Epidemiology"
subtitle: "A blended teaching workflow combining the mechanics of learning with a real WNV prediction problem"
author: "Federica Gazzelloni"
date: "2026-09-28"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

## Workshop notebook

Perspectives:

1.  a conceptual explanation of how machine learning models learn;
2.  a realistic epidemiology workflow for predicting West Nile virus (WNV) activity.

The goal is to understand both the mathematical idea of learning and the practical pipeline used in applied work.

By the end, we should be able to answer four questions:

1.  What is the model?
2.  What is the loss?
3.  What is the optimisation procedure?
4.  What is learned?

The central message is:

> Learning means using data to determine a model's parameters or structure. Optimisation is the procedure used to find useful values or structures according to an objective.

------------------------------------------------------------------------

# 1. Model, loss, optimisation, and learning

These terms are connected, but they are not interchangeable.

| Concept      | Question                                | Example               |
|------------------------|------------------------|------------------------|
| Model        | How are inputs turned into predictions? | $\hat y = wx + b$     |
| Loss         | How wrong are the predictions?          | Mean squared error    |
| Optimisation | How do we search for a better model?    | Gradient descent      |
| Learning     | What changes because of the data?       | Values of $w$ and $b$ |

The learning cycle is:

``` text
training data
     |
     v
make predictions with the current model
     |
     v
calculate the loss
     |
     v
use an optimisation rule to update the model
     |
     +-------------------- repeat for another epoch
```

Optimisation is therefore not an extra step after learning. In many models, it is the mechanism through which learning happens.

------------------------------------------------------------------------

# 2. A small linear-regression example

To make learning concrete, we begin with a tiny regression example.

```{r}
library(ggplot2)

theme_set(theme_minimal(base_size = 13))

training_data <- data.frame(
  x = 1:8,
  y = c(2.9, 3.4, 4.5, 5.0, 6.3, 6.7, 7.8, 8.1)
)

training_data
```

The model is:

$$
\hat y_i = wx_i + b,
$$

where:

- $w$ is the slope;
- $b$ is the intercept;
- $\hat y_i$ is the prediction for observation $i$.

At the start, both parameters are set to zero.

```{r}
w <- 0
b <- 0

training_data$initial_prediction <- w * training_data$x + b
training_data
```

------------------------------------------------------------------------

# 3. The loss function

We need a way to measure the gap between the observed values and the predictions. We use mean squared error:

$$
\operatorname{MSE}(w,b)
=
\frac{1}{n}\sum_{i=1}^{n}\left(\hat y_i-y_i\right)^2.
$$

```{r}
mse <- function(observed, predicted) {
  mean((predicted - observed)^2)
}

initial_loss <- mse(
  observed = training_data$y,
  predicted = training_data$initial_prediction
)

initial_loss
```

A loss value tells us how well the current parameters fit the training data. It does not update the parameters by itself.

------------------------------------------------------------------------

# 4. Batch gradient descent from scratch

Example from the article: [ml-in-epidemiology-article](https://arxiv.org/abs/2602.16352)

Gradient descent asks: in which direction should each parameter move to reduce the loss?

For MSE, the gradients are:

$$
\frac{\partial \operatorname{MSE}}{\partial w}
=
\frac{2}{n}\sum_{i=1}^{n}(\hat y_i-y_i)x_i
$$

and

$$
\frac{\partial \operatorname{MSE}}{\partial b}
=
\frac{2}{n}\sum_{i=1}^{n}(\hat y_i-y_i).
$$

The update rule is:

$$
w_{t+1}=w_t-\eta\frac{\partial \operatorname{MSE}}{\partial w},
\qquad
b_{t+1}=b_t-\eta\frac{\partial \operatorname{MSE}}{\partial b},
$$

where $\eta$ is the learning rate.

This is batch gradient descent because every update uses all observations.

```{r}
# Starting values
w <- 0
b <- 0

learning_rate <- 0.01
n_epochs <- 5000 # 1 epoch = 1 gradient update

loss_history <- data.frame(
  epoch = 0:n_epochs,
  loss = NA_real_,
  w = NA_real_,
  b = NA_real_
)

starting_prediction <- w * training_data$x + b
loss_history[1, ] <- c(
  epoch = 0,
  loss = mse(training_data$y, starting_prediction),
  w = w,
  b = b
)

for (epoch in 1:n_epochs) {
  prediction <- w * training_data$x + b
  error <- prediction - training_data$y

  gradient_w <- mean(2 * error * training_data$x)
  gradient_b <- mean(2 * error)

  w <- w - learning_rate * gradient_w
  b <- b - learning_rate * gradient_b

  updated_prediction <- w * training_data$x + b
  loss_history[epoch + 1, ] <- c(
    epoch = epoch,
    loss = mse(training_data$y, updated_prediction),
    w = w,
    b = b
  )
}

learned_w <- w
learned_b <- b

data.frame(
  parameter = c("intercept b", "slope w"),
  learned_value = c(learned_b, learned_w)
)
```

The code follows the repeated pattern:

``` text
predict -> calculate loss-related gradients -> update w and b -> repeat
```

This is the essence of learning.

------------------------------------------------------------------------

# 5. Compare the result with lm()

Base R can fit the same linear model directly.

```{r}
lm_fit <- lm(y ~ x, data = training_data)
lm_coefficients <- coef(lm_fit)

coefficient_comparison <- data.frame(
  parameter = c("intercept b", "slope w"),
  gradient_descent = c(learned_b, learned_w),
  lm = c(unname(lm_coefficients[1]), unname(lm_coefficients[2]))
)

coefficient_comparison$difference <-
  coefficient_comparison$gradient_descent - coefficient_comparison$lm

coefficient_comparison
```

```{r}
stopifnot(
  abs(learned_b - unname(lm_coefficients[1])) < 0.001,
  abs(learned_w - unname(lm_coefficients[2])) < 0.001,
  tail(loss_history$loss, 1) < loss_history$loss[1]
)
```

This shows that different numerical routes can produce the same fitted model.

------------------------------------------------------------------------

# 6. Why this matters for epidemiology

The same general ideas appear in real public-health prediction tasks.

The central question is:

> Can state-level surveillance information predict whether **West Nile virus** activity will be observed in the next four weeks?

The goal is not causation. The goal is to learn a pattern from historical data and check whether that pattern generalizes to new observations.

In supervised learning, the model learns a mapping from features $x$ to an outcome $y$:

$$
\hat{f}: X \rightarrow Y.
$$

- Features ($x$): recent reported cases, recent weeks active, region, and calendar timing.
- Outcome ($y$): whether WNV activity occurs in the next four weeks.
- Training data: historical data used to learn the pattern.
- Test data: unseen data used to evaluate generalization.

------------------------------------------------------------------------

# 7. WNV workflow: data, features, and prediction task

We now move from the abstract learning example to a real epidemiology workflow.

```{r}
# Basic libraries
library(ggplot2)
library(dplyr)
library(data.table)
library(maps)
library(pROC)
```

```{r}
wnv_path <- "data/processed/wnv_state_week.csv"
```

```{r}
# Read the processed WNV dataset
wnv <- read.csv(wnv_path, stringsAsFactors = FALSE)

# Keep useful columns
wnv <- wnv[wnv$week >= 18 & wnv$week <= 44, ]
wnv <- wnv[, c("state", "region", "year", "week",
               "cases_last4", "active_weeks_last8", "activity_next_4w")]

# Rename columns
names(wnv) <- c("state", "region", "year", "week",
                "recent_cases", "recent_active_weeks", "outcome")

# Prepare the target for classification
wnv$region <- factor(wnv$region)
wnv$outcome <- factor(wnv$outcome, levels = c("no", "yes"))

# Explore the class distribution
table(wnv$outcome)
round(prop.table(table(wnv$outcome)), 3)
```

We split the data into training and holdout test periods.

```{r}
# Split into training and test data
train <- subset(wnv, year <= 2024)
test <- subset(wnv, year == 2025)
```

------------------------------------------------------------------------

# 8. A simple logistic-regression model

This is a classic supervised learning setup: a model learns from historical state-week observations and predicts future activity.

```{r}
model <- glm(
  outcome ~ week + recent_cases + recent_active_weeks + region,
  data = train,
  family = binomial
)

summary(model)
```

```{r}
# Predict probabilities on unseen data
prob_yes <- predict(model, newdata = test, type = "response")

# Convert probabilities to class labels
predicted <- ifelse(prob_yes >= 0.5, "yes", "no")
predicted <- factor(predicted, levels = c("no", "yes"))

# Compare predictions with reality
confusion <- table(Actual = test$outcome, Predicted = predicted)
confusion

accuracy <- sum(diag(confusion)) / sum(confusion)
cat("Accuracy:", round(accuracy, 3), "\n")
```

This workflow shows the classic machine-learning cycle:

- learn on training data;
- predict on unseen data;
- compare predictions to observed outcomes;
- evaluate generalisation, not just fit.

------------------------------------------------------------------------

# 9. Model comparison in the modern workflow

The more modern workflow compares multiple learners and uses probabilities directly.

```{r}
library(mlr3)
library(mlr3learners)
```

```{r}
wnv_ml <- fread(wnv_path) |> # Similar to read.csv() and read.delim()
  filter(week >= 18, week <= 44) |>
  select(state, region, year, week,
         cases_lag1, cases_lag2, cases_lag4,
         cases_last4, cases_last8, active_weeks_last8,
         week_sin, week_cos, activity_next_4w) |>
  mutate(
    region = factor(region, levels = c("Northeast", "South", "North Central", "West")),
    activity_next_4w = factor(activity_next_4w, levels = c("no", "yes"))
  )

feature_columns <- c(
  "region",
  "cases_lag1",
  "cases_lag2",
  "cases_lag4",
  "cases_last4",
  "cases_last8",
  "active_weeks_last8",
  "week_sin",
  "week_cos"
)

train_data <- wnv_ml |> filter(year <= 2024)
test_data <- wnv_ml |> filter(year == 2025)

train_task <- TaskClassif$new(
  id = "wnv-train",
  backend = train_data[, c("activity_next_4w", feature_columns), with = FALSE],
  target = "activity_next_4w",
  positive = "yes"
)

test_task <- TaskClassif$new(
  id = "wnv-test",
  backend = test_data[, c("activity_next_4w", feature_columns), with = FALSE],
  target = "activity_next_4w",
  positive = "yes"
)

learner_logistic <- lrn(
  "classif.log_reg",
  id = "Logistic regression",
  predict_type = "prob"
)

learner_tree <- lrn(
  "classif.rpart",
  id = "Decision tree",
  predict_type = "prob",
  cp = 0.01,
  maxdepth = 5,
  minsplit = 40,
  xval = 0
)

learner_forest <- lrn(
  "classif.ranger",
  id = "Random forest",
  predict_type = "prob",
  num.trees = 500,
  mtry = 3,
  min.node.size = 10,
  importance = "permutation",
  num.threads = 1,
  seed = 20260927
)

learners <- list(
  logistic = learner_logistic,
  tree = learner_tree,
  forest = learner_forest
)

fitted <- lapply(learners, function(l) {
  l$clone(deep = TRUE)$train(train_task)
})

holdout_predictions <- rbindlist(lapply(names(fitted), function(model_name) {
  pred <- fitted[[model_name]]$predict(test_task)
  data.table(
    model = model_name,
    state = test_data$state,
    year = test_data$year,
    week = test_data$week,
    truth = pred$truth,
    probability = pred$prob[, "yes"]
  )
}))

summarise_model <- function(df, threshold = 0.5) {
  response <- ifelse(df$probability >= threshold, "yes", "no")
  response <- factor(response, levels = c("no", "yes"))
  truth <- factor(df$truth, levels = c("no", "yes"))

  tp <- sum(truth == "yes" & response == "yes")
  fn <- sum(truth == "yes" & response == "no")
  tn <- sum(truth == "no" & response == "no")
  fp <- sum(truth == "no" & response == "yes")

  sensitivity <- tp / (tp + fn)
  specificity <- tn / (tn + fp)
  accuracy <- (tp + tn) / length(truth)
  bacc <- mean(c(sensitivity, specificity))

  data.table(
    threshold = threshold,
    sensitivity = sensitivity,
    specificity = specificity,
    accuracy = accuracy,
    balanced_accuracy = bacc,
    tp = tp,
    fp = fp,
    tn = tn,
    fn = fn
  )
}

summary_table <- rbindlist(lapply(split(holdout_predictions, by = "model", keep.by = TRUE), function(x) {
  summarise_model(x, threshold = 0.5)
}), idcol = "model")

summary_table
```

This is framing of supervised learning and model evaluation: compare models, use probabilities, and evaluate generalization on unseen data.

------------------------------------------------------------------------

# 10. Visualization

### ROC curve

```{r}
roc_data <- rbindlist(lapply(unique(holdout_predictions$model), 
                             function(m) {
  pred <- holdout_predictions[model == m]
  obj <- roc(pred$truth, pred$probability, 
             levels = c("no", "yes"), quiet = TRUE)
  data.table(
    model = m,
    fpr = rev(obj$specificities),
    tpr = rev(obj$sensitivities)
  )
}))

ggplot(roc_data, aes(x = 1 - fpr, y = tpr, colour = model)) +
  geom_line(linewidth = 1) +
  geom_abline(intercept = 0, slope = 1, 
              linetype = 2, colour = "grey55") +
  labs(
    title = "ROC curves for WNV prediction models",
    x = "False-positive rate",
    y = "Sensitivity",
    colour = "Model"
  ) +
  theme_minimal()
```

### Confusion matrix plot

```{r}
confusion_data <- rbindlist(lapply(unique(holdout_predictions$model), function(m) {
  pred <- holdout_predictions[model == m]
  pred$response <- ifelse(pred$probability >= 0.5, "yes", "no")
  tab <- table(Actual = pred$truth, Predicted = pred$response)
  data.table(
    model = m,
    truth = rep(c("no", "yes"), each = 2),
    response = rep(c("no", "yes"), times = 2),
    count = as.vector(tab)
  )
}))

ggplot(confusion_data, aes(response, truth, fill = count)) +
  geom_tile(color = "white") +
  geom_text(aes(label = count), fontface = "bold") +
  scale_fill_gradient(low = "#F3E8FF", high = "#6D28D9") +
  facet_wrap(~ model) +
  labs(
    title = "Confusion matrices on the 2025 holdout set",
    x = "Predicted class",
    y = "Observed class",
    fill = "Count"
  ) +
  theme_minimal()
```

### Geographic map of predictions

```{r}
state_predictions <- holdout_predictions |> 
  group_by(state) |>
  summarise(predicted_probability = mean(probability), .groups = "drop")

state_map <- map_data("state")
state_map <- state_map |>
  rename(state = region) |>
  left_join(
    state_predictions |> mutate(state = tolower(state)),
    by = "state"
  )

ggplot(state_map, aes(long, lat, group = group, fill = predicted_probability)) +
  geom_polygon(color = "white", linewidth = 0.2) +
  coord_map() +
  scale_fill_gradient2(
    low = "#FDE68A",
    mid = "#F59E0B",
    high = "#B91C1C",
    midpoint = median(state_map$predicted_probability, na.rm = TRUE)
  ) +
  labs(
    title = "Predicted WNV activity probability by state",
    x = "",
    y = "",
    fill = "Predicted prob."
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```

------------------------------------------------------------------------

# 11. What the results mean

The point of the workflow is not only to fit a model but to explain how machine learning works in practice.

A typical sequence is:

- Use historical data to learn patterns.
- Choose a model and specify the outcome.
- Train the model on observed labels.
- Predict on unseen data.
- Evaluate the output with appropriate metrics.
- Interpret the model in context.

This is the bridge between the abstract learning cycle and applied epidemiology.

------------------------------------------------------------------------

# 12. Final takeaway

The conceptual notebook shows that learning is about using data to update model parameters or structure according to a loss function and an optimization scheme.

The epidemiology workflow shows how the same logic is used in a real decision problem: predict whether WNV activity will be observed in future weeks using information available at the current time.

The same essential pattern appears in both:

``` text
data -> model -> predictions -> loss -> update -> repeat -> evaluation
```

That is the heart of machine learning.
